Discovering governing reactions from concentration data

### 2018-08-27 docs/manuscript.tex
wip

\usepackage[left=2.5cm,right=2.6cm,top=3cm,bottom=3cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[intlimits]{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage[version=4]{mhchem}

@@

	For $\alpha=0$ the minimization problem reduces to constrained least-squares. In order to solve (\ref{method:minimizationproblem}) the sequential least-squares minimizer SLSQP \cite{Kraft1988} is applied via the software package SciPy \cite{SciPy}. Since only the concentration data $\mathbf{X}$ is available but not its temporal derivative $\dot{\mathbf{X}}$, it is approximated numerically by second order finite differences with the exception of boundary data.

@@

    To this end, let $A$, $B$, and $C$ be three species of proteins which are being translated from a corresponding mRNA molecule. Each mRNA in turn has a corresponding DNA which it is transcribed from. The proteins and mRNA molecules decay over time whereas the DNA concentration remains constant.

    For each of the three species one can formulate the following basic reactions
	\begin{align*}
        \ce{ DNA_i } & \ce{ -> DNA_i + mRNA_i } &\text{(transcription)}, \\
        \ce{ mRNA_i } & \ce{ -> mRNA_i + Protein_i } &\text{(translation)}, \\
        \ce{ mRNA_i } & \ce{ -> } \emptyset &\text{(decay of mRNA)}, \\
        \ce{ Protein_i } & \ce{ -> } \emptyset &\text{(decay of Protein)},
    \end{align*}
    for $i=A,B,C$.

    The protein concentrations of $A$, $B$, and $C$ regulate each other in a negative way by hindering the transcription process. In the law of mass action model we account for this by a reaction. A repression of species $j$ by species $i$ can be modeled as
    \begin{equation*}
        \ce{ Protein_i + mRNA_j -> Protein_i } \quad \text{(regulation by repression)}.
    \end{equation*}

    In our example proteins of type $A$ regulate the mRNA$_\mathrm{B}$ molecules, proteins of type $B$ regulate the mRNA$_\mathrm{A}$ molecules and proteins of type $C$ regulate the mRNA$_\mathrm{A}$ molecules. The network of all species and reactions is depicted in Figure \ref{fig:network}. This serves as reference model to generate the time series of concentrations.
    \begin{figure}
        \centering
        \includegraphics[width=.5\textwidth]{./figures_tex/scheme.pdf}
        \caption{The regulation network example described in Sec.~\ref{sec:generegulation}. \textbf{(a)}: Each circle depicts a species, each arrow corresponds to one reaction. Blue arrows denote transcription from DNA to mRNA, green arrows denote translation from mRNA to protein, and red arrows denote the regulatory network. \textbf{(b)}: Concentration timeseries generated from the reaction network shown in~(a). The initial condition prescribes positive concentration values only for B protein and $\mathrm{mRNA}_\mathrm{A}$ species. This initial condition is used in the subsequent sections for further analysis.}
        \label{fig:network}
    \end{figure}

@@

				$\ce{DNA_A}$ &$\ce{->}$& $\ce{DNA_A + mRNA_A }$ & $k_1 = 1.8$ & transcription of $\ce{mRNA_A}$\\
				$\ce{mRNA_A}$ &$\ce{->}$& $\ce{mRNA_A + A}$ & $k_2 = 2.1$ & translation of $\ce{A}$ proteins\\
				$\ce{mRNA_A}$ &$\ce{->}$& $\ce{\emptyset}$ & $k_3 = 1.3$ & $\ce{mRNA_A}$ decay\\
				$\ce{A}$ &$\ce{->}$& $\ce{\emptyset }$ & $k_4 = 1.5$ & decay of $\ce{A}$ proteins\\
				$\ce{DNA_B}$ &$\ce{->}$& $\ce{DNA_B + mRNA_B }$ & $k_5 = 2.2$ & transcription of $\ce{mRNA_B}$\\
				$\ce{mRNA_B}$ &$\ce{->}$& $\ce{mRNA_B + B}$ & $k_6 = 2.0$ & translation of $\ce{B}$ proteins\\
				$\ce{mRNA_B}$ &$\ce{->}$& $\ce{\emptyset }$ & $k_7 = 2.0$ & $\ce{mRNA_B}$ decay\\
				$\ce{B}$ &$\ce{->}$& $\ce{\emptyset }$ & $k_8 = 2.5$ & decay of $\ce{B}$ proteins\\
				$\ce{DNA_C}$ &$\ce{->}$& $\ce{DNA_C + mRNA_C }$ & $k_9 = 3.2$ & transcription of $\ce{mRNA_C}$\\
				$\ce{mRNA_C}$ &$\ce{->}$& $\ce{mRNA_C + C}$ & $k_{10} = 3.0$ & translation of $\ce{C}$ proteins\\
				$\ce{mRNA_C}$ &$\ce{->}$& $\ce{\emptyset }$ & $k_{11} = 2.3$ & $\ce{mRNA_C}$ decay\\
				$\ce{C}$ &$\ce{->}$& $\ce{\emptyset }$ & $k_{12} = 2.5$ & decay of $\ce{C}$ proteins\\
				$\ce{mRNA_A + A}$ &$\ce{->}$& $A$ & $k_{13} = 0$ & self regulation of $\ce{A}$ proteins\\
				$\ce{mRNA_B + B}$ &$\ce{->}$& $B$ & $k_{14} = 0$ & self regulation of $\ce{B}$ proteins\\
				$\ce{mRNA_C + C}$ &$\ce{->}$& $C$ & $k_{15} = 0$ & self regulation of $\ce{C}$ proteins\\
				$\ce{mRNA_B + A}$ &$\ce{->}$& $A$ & $k_{16} = 0$ & cyclic regulation of $\ce{A}$ proteins\\
				$\ce{mRNA_C + B}$ &$\ce{->}$& $B$ & $k_{17} = 0$ & cyclic regulation of $\ce{B}$ proteins\\
				$\ce{mRNA_A + C}$ &$\ce{->}$& $C$ & $k_{18} = 0$ & cyclic regulation of $\ce{C}$ proteins\\
				$\ce{mRNA_C + A}$ &$\ce{->}$& $A$ & $k_{16} = 6.0$ & cyclic regulation of $\ce{A}$ proteins\\
				$\ce{mRNA_B + C}$ &$\ce{->}$& $C$ & $k_{17} = 4.0$ & cyclic regulation of $\ce{C}$ proteins\\
				$\ce{mRNA_A + B}$ &$\ce{->}$& $B$ & $k_{18} = 3.0$ & cyclic regulation of $\ce{B}$ proteins\\
				$\ce{mRNA_A + A}$ &$\ce{->}$& $\ce{mRNA_A}$ & $k_{19} = 0$ & artificial fusion\\
				$\ce{mRNA_B + B}$ &$\ce{->}$& $\ce{mRNA_B}$ & $k_{20} = 0$ & artificial fusion\\
				$\ce{mRNA_C + C}$ &$\ce{->}$& $\ce{mRNA_C}$ & $k_{21} = 0$ & artificial fusion\\
				$\ce{mRNA_A + B}$ &$\ce{->}$& $\ce{mRNA_A}$ & $k_{22} = 0$ & artificial fusion\\
				$\ce{mRNA_B + C}$ &$\ce{->}$& $\ce{mRNA_B}$ & $k_{23} = 0$ & artificial fusion\\
				$\ce{mRNA_C + A}$ &$\ce{->}$& $\ce{mRNA_C}$ & $k_{24} = 0$ & artificial fusion\\
				$\ce{mRNA_A + C}$ &$\ce{->}$& $\ce{mRNA_A}$ & $k_{25} = 0$ & artificial fusion\\
				$\ce{mRNA_B + A}$ &$\ce{->}$& $\ce{mRNA_B}$ & $k_{26} = 0$ & artificial fusion\\
				$\ce{mRNA_C + B}$ &$\ce{->}$& $\ce{mRNA_C}$ & $k_{27} = 0$ & artificial fusion\\
				$\ce{A + A}$ &$\ce{->}$& $\ce{A}$ & $k_{28} = 0$ & $\ce{A}$ regulates $\ce{A}$\\
				$\ce{B + B}$ &$\ce{->}$& $\ce{B}$ & $k_{29} = 0$ & $\ce{B}$ regulates $\ce{B}$\\
				$\ce{C + C}$ &$\ce{->}$& $\ce{C}$ & $k_{30} = 0$ & $\ce{C}$ regulates $\ce{C}$\\
				$\ce{B + A}$ &$\ce{->}$& $\ce{A}$ & $k_{31} = 0$ & artificial fusion between proteins\\
				$\ce{C + B}$ &$\ce{->}$& $\ce{B}$ & $k_{32} = 0$ & artificial fusion between proteins\\
				$\ce{A + C}$ &$\ce{->}$& $\ce{C}$ & $k_{33} = 0$ & artificial fusion between proteins\\
				$\ce{C + A}$ &$\ce{->}$& $\ce{A}$ & $k_{34} = 0$ & artificial fusion between proteins\\
				$\ce{B + C}$ &$\ce{->}$& $\ce{C}$ & $k_{35} = 0$ & artificial fusion between proteins\\
				$\ce{A + B}$ &$\ce{->}$& $\ce{B}$ & $k_{36} = 0$ & artificial fusion between proteins\\
				$\ce{A}$ &$\ce{->}$& $\ce{B}$ & $k_{37} = 0$ & artificial conversion between proteins\\
				$\ce{B}$ &$\ce{->}$& $\ce{C}$ & $k_{38} = 0$ & artificial conversion between proteins\\
				$\ce{C}$ &$\ce{->}$& $\ce{A}$ & $k_{39} = 0$ & artificial conversion between proteins\\
				$\ce{A}$ &$\ce{->}$& $\ce{C}$ & $k_{40} = 0$ & artificial conversion between proteins\\
				$\ce{C}$ &$\ce{->}$& $\ce{B}$ & $k_{41} = 0$ & artificial conversion between proteins\\
				$\ce{B}$ &$\ce{->}$& $\ce{A}$ & $k_{42} = 0$ & artificial conversion between proteins

>>>>

\usepackage[left=2.5cm,right=2.6cm,top=3cm,bottom=3cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[intlimits]{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{xcolor}
\usepackage{subcaption}

@@

	For $\alpha=0$ the minimization problem reduces to constrained least-squares. In order to solve (\ref{method:minimizationproblem}) the numerical sequential least-squares minimizer SLSQP \cite{Kraft1988} is applied via the software package SciPy \cite{SciPy}. Since only the concentration data $\mathbf{X}$ is available but not its temporal derivative $\dot{\mathbf{X}}$, it is approximated numerically by second order finite differences with the exception of boundary data.

@@

    To this end, let $\mathrm{A}$, $\mathrm{B}$, and $\mathrm{C}$ be three species of proteins which are being translated from a corresponding $\mathrm{mRNA}$ molecule. Each $\mathrm{mRNA}$ in turn has a corresponding $\mathrm{DNA}$ which it is transcribed from. The proteins and $\mathrm{mRNA}$ molecules decay over time whereas the $\mathrm{DNA}$ concentration remains constant.

    The following basic reactions can be formulated
	\begin{align*}
		\mathrm{DNA}_i &\rightharpoonup \mathrm{DNA}_i + \mathrm{mRNA}_i &&\text{(transcription)},\\
		\mathrm{mRNA}_i &\rightharpoonup \mathrm{mRNA}_i + \mathrm{Protein}_i &&\text{(translation)},\\
		\mathrm{mRNA}_i &\rightharpoonup \emptyset &&\text{(decay of mRNA)},\\
		\mathrm{Protein}_i &\rightharpoonup \emptyset &&\text{(decay of Protein)},
    \end{align*}
    for each of the species $i\in S:=\{\mathrm{A},\mathrm{B},\mathrm{C}\}$.

    The protein concentrations of $\mathrm{A}$, $\mathrm{B}$, and $\mathrm{C}$ regulate each other in a negative way by hindering the transcription process. In the law of mass action model we account for this by a reaction. A repression of species $j\in S$ by species $i\in S$ can be modeled as
    \begin{align*}
    i + \mathrm{mRNA}_j \rightharpoonup i &&&\text{(regulation by repression).}
    \end{align*}

    In our example proteins of type $\mathrm{A}$ regulate the $\mathrm{mRNA}_\mathrm{B}$ molecules, proteins of type $\mathrm{B}$ regulate the $\mathrm{mRNA}_\mathrm{A}$ molecules and proteins of type $\mathrm{C}$ regulate the $\mathrm{mRNA}_\mathrm{A}$ molecules. The network of all species and reactions is depicted in Figure \ref{fig:network}. This serves as reference model to generate the time series of concentrations.
    \begin{figure}
        \centering
        \includegraphics[width=.5\textwidth]{./figures_tex/scheme.pdf}
        \caption{The regulation network example described in Sec.~\ref{sec:generegulation}. \textbf{(a)}: Each circle depicts a species, each arrow corresponds to one reaction. Blue arrows denote transcription from DNA to mRNA, green arrows denote translation from mRNA to protein, and red arrows denote the regulatory network. \textbf{(b)}: Concentration timeseries generated from the reaction network shown in~(a). The initial condition prescribes positive concentration values only for B protein and $\mathrm{mRNA}_\mathrm{A}$ species. This initial condition is used in the subsequent sections for further analysis.}
        \label{fig:network}
    \end{figure}

@@

				$\mathrm{DNA}_\mathrm{A}$ &$\rightharpoonup$& $\mathrm{DNA}_\mathrm{A} + \mathrm{mRNA}_\mathrm{A}$ & $k_1 = 1.8$ & transcription of $\mathrm{mRNA}_\mathrm{A}$\\
				$\mathrm{mRNA}_\mathrm{A}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{A} + \mathrm{A}$ & $k_2 = 2.1$ & translation of $\mathrm{A}$ proteins\\
				$\mathrm{mRNA}_\mathrm{A}$ &$\rightharpoonup$& $\emptyset$ & $k_3 = 1.3$ & $\mathrm{mRNA}_\mathrm{A}$ decay\\
				$\mathrm{A}$ &$\rightharpoonup$& $\emptyset$ & $k_4 = 1.5$ & decay of $\mathrm{A}$ proteins\\
				$\mathrm{DNA}_\mathrm{B}$ &$\rightharpoonup$& $\mathrm{DNA}_\mathrm{B} + \mathrm{mRNA}_\mathrm{B}$ & $k_5 = 2.2$ & transcription of $\mathrm{mRNA}_\mathrm{B}$\\
				$\mathrm{mRNA}_\mathrm{B}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{B} + \mathrm{B}$ & $k_6 = 2.0$ & translation of $\mathrm{B}$ proteins\\
				$\mathrm{mRNA}_\mathrm{B}$ &$\rightharpoonup$& $\emptyset$ & $k_7 = 2.0$ & $\mathrm{mRNA}_\mathrm{B}$ decay\\
				$\mathrm{B}$ &$\rightharpoonup$& $\emptyset$ & $k_8 = 2.5$ & decay of $\mathrm{B}$ proteins\\
				$\mathrm{DNA}_\mathrm{C}$ &$\rightharpoonup$& $\mathrm{DNA}_\mathrm{C} + \mathrm{mRNA}_\mathrm{C}$ & $k_9 = 3.2$ & transcription of $\mathrm{mRNA}_\mathrm{C}$\\
				$\mathrm{mRNA}_\mathrm{C}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{C} + \mathrm{C}$ & $k_{10} = 3.0$ & translation of $\mathrm{C}$ proteins\\
				$\mathrm{mRNA}_\mathrm{C}$ &$\rightharpoonup$& $\emptyset$ & $k_{11} = 2.3$ & $\mathrm{mRNA}_\mathrm{C}$ decay\\
				$\mathrm{C}$ &$\rightharpoonup$& $\emptyset$ & $k_{12} = 2.5$ & decay of $\mathrm{C}$ proteins\\
				$\mathrm{mRNA}_\mathrm{A} + \mathrm{A}$ &$\rightharpoonup$& $\mathrm{A}$ & $k_{13} = 0$ & self regulation of $\mathrm{A}$ proteins\\
				$\mathrm{mRNA}_\mathrm{B} + \mathrm{B}$ &$\rightharpoonup$& $\mathrm{B}$ & $k_{14} = 0$ & self regulation of $\mathrm{B}$ proteins\\
				$\mathrm{mRNA}_\mathrm{C} + \mathrm{C}$ &$\rightharpoonup$& $\mathrm{C}$ & $k_{15} = 0$ & self regulation of $\mathrm{C}$ proteins\\
				$\mathrm{mRNA}_\mathrm{B} + \mathrm{A}$ &$\rightharpoonup$& $\mathrm{A}$ & $k_{16} = 0$ & cyclic regulation of $\mathrm{A}$ proteins\\
				$\mathrm{mRNA}_\mathrm{C} + \mathrm{B}$ &$\rightharpoonup$& $\mathrm{B}$ & $k_{17} = 0$ & cyclic regulation of $\mathrm{B}$ proteins\\
				$\mathrm{mRNA}_\mathrm{A} + \mathrm{C}$ &$\rightharpoonup$& $\mathrm{C}$ & $k_{18} = 0$ & cyclic regulation of $\mathrm{C}$ proteins\\
				$\mathrm{mRNA}_\mathrm{C} + \mathrm{A}$ &$\rightharpoonup$& $\mathrm{A}$ & $k_{16} = 6.0$ & cyclic regulation of $\mathrm{A}$ proteins\\
				$\mathrm{mRNA}_\mathrm{B} + \mathrm{C}$ &$\rightharpoonup$& $\mathrm{C}$ & $k_{17} = 4.0$ & cyclic regulation of $\mathrm{C}$ proteins\\
				$\mathrm{mRNA}_\mathrm{A} + \mathrm{B}$ &$\rightharpoonup$& $\mathrm{B}$ & $k_{18} = 3.0$ & cyclic regulation of $\mathrm{B}$ proteins\\
				$\mathrm{mRNA}_\mathrm{A} + \mathrm{A}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{A}$ & $k_{19} = 0$ & artificial fusion\\
				$\mathrm{mRNA}_\mathrm{B} + \mathrm{B}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{B}$ & $k_{20} = 0$ & artificial fusion\\
				$\mathrm{mRNA}_\mathrm{A} + \mathrm{B}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{A}$ & $k_{21} = 0$ & artificial fusion\\
				$\mathrm{mRNA}_\mathrm{B} + \mathrm{C}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{B}$ & $k_{22} = 0$ & artificial fusion\\
				$\mathrm{mRNA}_\mathrm{C} + \mathrm{A}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{C}$ & $k_{23} = 0$ & artificial fusion\\
				$\mathrm{mRNA}_\mathrm{A} + \mathrm{C}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{A}$ & $k_{24} = 0$ & artificial fusion\\
				$\mathrm{mRNA}_\mathrm{B} + \mathrm{A}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{B}$ & $k_{25} = 0$ & artificial fusion\\
				$\mathrm{A} + \mathrm{A}$ &$\rightharpoonup$& $\mathrm{A}$ & $k_{26} = 0$ & $\mathrm{A}$ regulates $\mathrm{A}$\\
				$\mathrm{B} + \mathrm{B}$ &$\rightharpoonup$& $\mathrm{B}$ & $k_{27} = 0$ & $\mathrm{B}$ regulates $\mathrm{B}$\\
				$\mathrm{C} + \mathrm{C}$ &$\rightharpoonup$& $\mathrm{C}$ & $k_{28} = 0$ & $\mathrm{C}$ regulates $\mathrm{C}$\\
				$\mathrm{B} + \mathrm{A}$ &$\rightharpoonup$& $\mathrm{A}$ & $k_{29} = 0$ & artificial fusion between proteins\\
				$\mathrm{C} + \mathrm{B}$ &$\rightharpoonup$& $\mathrm{B}$ & $k_{30} = 0$ & artificial fusion between proteins\\
				$\mathrm{A} + \mathrm{C}$ &$\rightharpoonup$& $\mathrm{C}$ & $k_{31} = 0$ & artificial fusion between proteins\\
				$\mathrm{C} + \mathrm{A}$ &$\rightharpoonup$& $\mathrm{A}$ & $k_{32} = 0$ & artificial fusion between proteins\\
				$\mathrm{B} + \mathrm{C}$ &$\rightharpoonup$& $\mathrm{C}$ & $k_{33} = 0$ & artificial fusion between proteins\\
				$\mathrm{A} + \mathrm{B}$ &$\rightharpoonup$& $\mathrm{B}$ & $k_{34} = 0$ & artificial fusion between proteins\\
				$\mathrm{A}$ &$\rightharpoonup$& $\mathrm{B}$ & $k_{35} = 0$ & artificial conversion between proteins\\
				$\mathrm{B}$ &$\rightharpoonup$& $\mathrm{C}$ & $k_{36} = 0$ & artificial conversion between proteins\\
				$\mathrm{C}$ &$\rightharpoonup$& $\mathrm{A}$ & $k_{37} = 0$ & artificial conversion between proteins\\
				$\mathrm{A}$ &$\rightharpoonup$& $\mathrm{C}$ & $k_{38} = 0$ & artificial conversion between proteins\\
				$\mathrm{C}$ &$\rightharpoonup$& $\mathrm{B}$ & $k_{39} = 0$ & artificial conversion between proteins\\
				$\mathrm{B}$ &$\rightharpoonup$& $\mathrm{A}$ & $k_{40} = 0$ & artificial conversion between proteins\\
				$\mathrm{mRNA}_\mathrm{B} + \mathrm{mRNA}_\mathrm{C}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{A}$ & $k_{41} = 0$ & artificial fusion\\
				$\mathrm{mRNA}_\mathrm{C} + \mathrm{mRNA}_\mathrm{B}$ &$\rightharpoonup$& $\mathrm{mRNA}_\mathrm{C}$ & $k_{42} = 0$ & artificial fusion\\
				$\mathrm{mRNA}_\mathrm{C} + \mathrm{A}$ &$\rightharpoonup$& $\mathrm{C}$ & $k_{43} = 0$ & artificial fusion\\
				$\mathrm{mRNA}_\mathrm{C} + \mathrm{C}$ &$\rightharpoonup$& $\mathrm{C}$ & $k_{43} = 0$ & artificial fusion
